### 2023-09-02 manuscript/submission2/outbreak_response_models_review_v2.tex

\subsection*{Locations studied}
Overall, most of the papers were about actual geographic locations $(78.9\%; 195/247)$ as compared to hypothetical locations $(21.1\%; 52/247)$. Among the former, some studied a geographic location spanning more than one country but not classifiable as a continent $(9.7\%; 19/195)$. Moreover, among those $19$ studies, West Africa was the most studied $(47.4\%; 9/19)$, followed by the whole globe $(36.8\%; 7/19)$, Southeast Asia $(10.6\%; 2/19)$, and the Northern Hemisphere $(5.3\%; 1/19)$.

@@

When aggregated into continents, the Americas were studied the most $(36.4\%; 68/187)$, followed by Asia $(25.1\%;47/187)$, Africa $(24.6\%; 46/187)$, Europe $(13.4\%; 25/187)$, and Oceania $(0.5\%; 1/187)$.

When disaggregated into countries, the United States was the most studied $(21.6\%; 38/176)$, followed by China $(10.8\%; 19/176)$, Canada $(6.2\%; 11/176)$, and Sierra Leone $(5.7\%; 10/176)$.

@@

\subsection*{Connection of authors to the location studied}
We investigated the connection of the authors to the location studied and found that, overall, there were more studies with at least one author connected to the studied location $(75.2\%; 118/157)$ than not $(24.8\%; 39/157)$.

When stratified by collaboration types, mixed collaborations were more likely to have studies with at least one author in the location studied $(84.0\%; 63/75)$ compared with the purely academic collaborations $(67.1\%; 55/82)$.

\subsection*{Diseases studied}
Overall, the number of studies per disease was disproportionately distributed (Table \ref{studies_per_disease_and_collaboration_type}). Influenza was the most studied $(57.4\%; 135/235)$, followed by Ebola $(14.5\%; 34/235)$, Dengue $(5.1\%; 12/236)$, and a tie between Cholera $(4.7\%; 11/235)$, and Measles $(4.7\%; 11/235)$.

When the diseases were broken down in terms of which collaboration types studied them, there were clear differences in their distribution. Influenza, Dengue, and Measles were more studied by mixed collaborations whereas Ebola and Cholera were more studied by purely academic collaborations.

\begin{table}[!h]
	\setlength\arrayrulewidth{1pt} 
	\centering
	\caption{\bf Number of studies per disease and collaboration type. Percentages are calculated from the row totals. The number of studies making up the percentages are shown in brackets. In counting the number of papers per disease studied, if a study was about more than one disease, we treated each instance as unique and counted them separately, leading to a total greater than the reported number of studies.}
	\begin{tabular}{|l c c c|}
		\hline
		\textbf{Disease}         & \textbf{Purely academic} & \textbf{Mixed} & \textbf{Total} \\ \hline
		Influenza                & 54.1\% (73)     & 45.9\% (62) & 135   \\ \hline
		Ebola                    & 70.6\% (24)     & 29.4\% (10) & 34    \\ \hline
		Dengue                   & 50.0\% (6)      & 50.0\% (6)  & 12    \\ \hline
		Cholera                  & 81.8\% (9)      & 18.2\% (2)  & 11    \\ \hline
		Measles                  & 36.4\% (4)      & 63.6\% (7)  & 11    \\ \hline
		Poliomyelitis            & 28.6\% (2)      & 71.4\% (5)  & 7     \\ \hline
		Tuberculosis             & 83.3\% (5)      & 16.7\% (1)  & 6     \\ \hline
		Varicella                & 50.0\% (2)      & 50.0\% (2)  & 4     \\ \hline
		Meningococcal meningitis & 33.3\% (1)      & 66.7\% (2)  & 3     \\ \hline
		Pertussis                & 50.0\% (1)      & 50.0\% (1)  & 2     \\ \hline
		Pneumococcal disease     & 50.0\% (1)      & 50.0\% (1)  & 2     \\ \hline
		Yellow fever             & 50.0\% (1)      & 50.0\% (1)  & 2     \\ \hline
		Hepatitis A             & 100.0\% (1)     & 0.0\% (0)   & 1     \\ \hline
		Rubella                  & 100.0\% (1)     & 0.0\% (0)   & 1     \\ \hline
		Typhoid                  & 100.0\% (1)     & 0.0\% (0)   & 1     \\ \hline
		Hepatitis B              & 0.0\% (0)       & 100.0\% (1) & 1     \\ \hline
		Malaria                  & 0.0\% (0)       & 100.0\% (1) & 1     \\ \hline
		Mumps                    & 0.0\% (0)       & 100.0\% (1) & 1    \\ \hline \rowcolor{gray!20}
		\textbf{Overall}			&	\textbf{56.2\% (132)}   			& \textbf{43.8\% (103)}   & \textbf{100\% (235)} \\
		\hline
	\end{tabular}
	\label{studies_per_disease_and_collaboration_type}
\end{table}

\subsection*{Intervention types}
There were more papers about non-vaccination interventions $(47.6\%; 108/227)$, followed by those that modelled vaccination in combination with other interventions or do-nothing counterfactuals $(41.0\%; 93/227)$, and those that modelled vaccination as a single intervention for side-by-side comparison with other non-vaccination interventions $(11.5\%; 26/227)$.

We also tabulated the top 2 interventions studied per disease (Table \ref{top2_interventions_per_disease}).

\begin{table}
	\setlength\arrayrulewidth{1pt} 
	\centering
	\caption{\bf Top 2 most studied interventions per disease. We counted the unique number of times an intervention was studied per disease and reported the top two.}
		\begin{tabular}{| l | l |}
			\hline
			\textbf{Disease} & \textbf{Intervention modelled} \\ \hline
			Cholera & vaccination \\ \hline
			Cholera & hygiene \\ \hline
			Dengue & vaccination \\ \hline
			Dengue & larvicides \\ \hline
			Ebola & isolation \\ \hline
			Ebola & safe burial \\ \hline
			Hepatitis a & vaccination \\ \hline
			Hepatitis b & vaccination \\ \hline
			Influenza & vaccination \\ \hline
			Influenza & school closure \\ \hline
			Malaria & drugs \\ \hline
			Malaria & treatment \\ \hline
			Measles & vaccination \\ \hline
			Measles & behavioural change \\ \hline
			Meningococcal meningitis & vaccination \\ \hline
			Meningococcal meningitis & PPE \\ \hline
			Pertussis & vaccination \\ \hline
			Pertussis & contact tracing \\ \hline
			Pneumococcal disease & treatment \\ \hline
			Poliomyelitis & vaccination \\ \hline
			Poliomyelitis & school closure \\ \hline
			Rubella & vaccination \\ \hline
			Tuberculosis & treatment \\ \hline
			Tuberculosis & vaccination \\ \hline
			Typhoid & screening \\ \hline
			Varicella & vaccination \\ \hline
			Varicella & isolation \\ \hline
			Yellow fever & vaccination \\ \hline
		\end{tabular}
			\label{top2_interventions_per_disease}
	\end{table}

\subsection*{Model structure, spatial heterogeneity, and model dynamics}
There were more compartmental models $(78.9\%; 179/227)$ than agent-based models $(19.8\%; 45/227)$ with no clear difference in preference between the two collaboration types.

Approximately a third of the papers included spatial heterogeneity $(28.6\%; 65/227)$ with no clear difference between the two collaboration types.

Deterministic models were the most common $(63.4\%; 144/227)$ compared to stochastic $(30.4\%; 69/227)$ and hybrid models $(6.2\%; 14/227)$, that is models with both deterministic and stochastic components. Mixed collaborations were more likely to use stochastic models $(39.8\%; 39/100)$ than purely academic collaborations $(23.3\%; 30/128)$.

\subsection*{Outcomes measured}
We counted the unique number of times an outcome was studied by each collaboration type (Table \ref{top 6 outcomes studied by collab types}). Final epidemic size was the top outcome of interest in both groups. The top 6 outcomes were common between both groups except for intervention cost, which was more of interest to purely academic collaborations but not to mixed collaborations. On the other hand, number of hospitalizations was more of interest to mixed collaborations than purely academic collaborations.

\begin{table}[h!]
	 \caption{\textbf{Top 6 outcomes studied per collaboration type. Percentages are calculated from the total of studies within each collaboration type. Actual numbers are provided in brackets.}}
	 \centering
	 \begin{tabular}{| l | l | l |}\hline
	 	\textbf{Collaboration type} & \textbf{Outcome measured} & \textbf{number of studies} \\ \hline
	 	Purely academic & final epidemic size & 22.0\%  (55)\\ \hline
	 	 & attack rate & 11.6\%  (29)\\ \hline
	 	 & timing of peak & 8.0\%  (20)\\ \hline
	 	 & cost & 7.2\%  (18)\\ \hline
	 	 & outbreak duration and timing & 6.0\%  (15)\\ \hline
	 	 & cases averted & 4.8\%  (12)\\ \hline
	 	Mixed & final epidemic size & 14.4\%  (31)\\ \hline
	 	 & attack rate & 13.4\%  (29)\\ \hline
	 	 & cases averted & 10.2\%  (22)\\ \hline
	 	 & timing of peak & 7.,4\%  (16)\\ \hline
	 	 & outbreak duration and timing & 6.0\%  (13)\\ \hline
	 	 & hospitalizations & 4.2\%   (9)\\ \hline
 	\end{tabular}
 	\label{top 6 outcomes studied by collab types}
 	\end{table}

\subsection*{Model parametrization and validation}
The top three most commonly used parametrization methods, that is, how model parameter values were obtained, included: combining literature sources and expert opinion/assumptions, literature sources and fitting to data, and literature sources only (Table \ref{model_parametrization_results}).

The most common parametrization method among mixed collaborations was the combination of literature sources and fitting $(28.6 \%; 28/98)$, whereas the combination of literature and expert opinion was the most common among purely academic collaborations $(27.1\%; 35/129)$.

\begin{table}[!h]
\setlength\arrayrulewidth{1pt} 
\centering
\caption{\bf Model parametrization methods. We define model parametrization as the method of determining the parameter values for the model. In the table, ``Literature'' means the model's parameters were obtained from literature sources. ``Expert opinion'' means the values were assumed in consultation with experts of the field. ``Fitted'' means the model’s parameters were obtained through some form of mathematical or statistical fitting to a time series of data. }
\begin{adjustwidth}{-2.10in}{0in}
\begin{tabular}{| p{0.17\textwidth}  p{0.14\textwidth}  p{0.14\textwidth}  p{0.12\textwidth}  p{0.13\textwidth}  p{0.13\textwidth}  p{0.12\textwidth}  p{0.1\textwidth}  p{0.07\textwidth} |} \hline 
\textbf{Collaboration type} & \textbf{Literature and expert opinion} & \textbf{Literature and fitted} & \textbf{Literature} &  \textbf{Literature, expert opinion, and fitted} & \textbf{Expert opinion} & \textbf{Fitted} & \textbf{Expert opinion and fitted} & \textbf{Total} \\ \hline
Academic & 27.1\% (35) & 21.7\% (28) & 17.8\% (23) & 10.1\% (13) & 14.0\% (18)  & 7.0\% (9) & 2.3\% (3) & 129 \\ \hline
Mixed & 24.5\% (24) & 28.6\% (28) & 14.3\% (14) & 19.4\% (19) & 7.1\% (7) & 5.1\% (5) & 1.0\% (1) & 98 \\ \hline \rowcolor{gray!20}
Overall & 26.3\% (59) & 24.7\% (56) & 16.3\% (37) & 14.1\% (32) & 11.0\% (25) & 6.2\% (14) & 1.8\% (4) & 227 \\ \hline 
\end{tabular}
\end{adjustwidth}
\label{model_parametrization_results}
\end{table}

We defined model validation as the method by which the model's performance was measured. Four categories of model validation were established for the purpose of this review. In Table \ref{model_validation_results}, we present the results. If the model was not validated with observed data or another model's output, we classified its validation as ``none''.  ``Data'' means the model's output was compared to independently observed data. ``Another model’s output'' means the model's output was compared to an independent model's output. Some studies also used data and another model's output. We classified these as ``data and another model".

Most studies were classified as none $(63.2\%; 144/227)$. Approximately a third used data to validate their models $(34.6\%; 79/227)$ (Table \ref{model_validation_results}).

In terms of differences between the collaboration types, mixed collaborations were more likely to validate their model with data or the output of an independent model.

\begin{table}[!h]
	\centering
	\setlength\arrayrulewidth{1pt} 
	\caption{\bf Model validation methods.}
	\begin{adjustwidth}{-0.00in}{0in}
		\begin{tabular}{| p{0.15\textwidth}  p{0.13\textwidth} p{0.13\textwidth} p{0.13\textwidth} p{0.13\textwidth} p{0.12\textwidth}|}
			\hline
			\textbf{Collaboration type} & \textbf{None} & \textbf{Data} & \textbf{Another model's output} & \textbf{Data and another model} & \textbf{Total} \\ \hline
			Academic & 71.3\% (92) & 27.9\% (36) & 0.8\% (1) & 0.0\% (0) & 129 \\ \hline
			Mixed & 53.1\% (52) & 42.9\% (42) & 3.1\% (3) & 1.0\% (1) & 98 \\ \hline \rowcolor{gray!20}
			Overall & 63.4\% (144) & 34.4\% (79) & 1.8\% (4) & 0.4\% (1) & 227\\ \hline
		\end{tabular}
	\end{adjustwidth}
	\label{model_validation_results}
\end{table}

\subsection*{Data and model simulation code}
More than half of the papers used datasets collected independent of the study for either the model parametrization or validation process $(55.5\%; 126/227)$. These papers were split equally between the two collaboration types.

The use of accessible data was, however, different between the two groups. Purely academic collaborations were more likely to use data that could be accessed in the public domain $(82.5\%; 52/63)$ compared to mixed collaborations $(58.7\%; 37/63)$.

Few papers provided access to the model simulation code $(1.7\%; 4/227)$. Here, if a paper indicated that the authors could be contacted for the code, it was deemed as inaccessible due to the many hurdles with getting the code in time. Additionally, some papers reported the computer application/software/program/package used (R, Python, C++, Matlab, and so forth) but we did not collect that information.

\subsection*{Patterns in the foot and mouth disease (FMD) literature}
We included $25$ foot and mouth disease modelling studies for comparison with the human disease literature. Due to the small number of studies, the absolute differences between the two collaboration types were generally not large enough to be considered (\nameref{S2_File}), hence, we report on the aggregated patterns.

Slightly more than half of the $25$ studies were authored by mixed collaborations $(56.0\%; 14/25)$. In terms of how connected the authors were to the locations studied, almost all the papers had at least one author affiliated to an institution in the location studied $(92.0\%; 23/25)$.

Overall, vaccination was modelled as a single intervention for comparison in more than half of the studies $(56.0\%; 14/25)$, followed by vaccination in combination with other interventions $(28.0\%; 7/25)$, and no vaccination $(16.0\%; 4/25)$.

@@

In general, the FMD models had more agent-based $(76.0\%; 19/25)$ and spatially explicit structure $(72.0\%; 18/25)$, largely used stochastic dynamics $(56.0\%; 14/25)$, and performed more sensitivity analyses $(56.0\%; 14/25)$.

Outbreak duration and timing, final epidemic size, and intervention costs were the top three outcomes of interest overall.

In terms of model parametrization, the most common method was the combined use of literature sources and expert opinion $(28.0\%; 7/25)$, followed by fitting to data only $(20.0\%; 5/25)$, and the combined use of literature sources, expert opinion, and fitting to data $(16.0\%; 4/25)$. In terms of model validation, most of the FMD models were not compared to observed data or the output of independent models $(72.0\%; 18/25)$.

\section*{Discussion}
For the period $1970-2019$, the outbreak response intervention impact modelling literature for human vaccine-preventable diseases was dominated by purely academic collaborations. This collaboration type, which we defined as papers with all authors affiliated to academic institutions, formed over $56\%$ of the literature. Both purely academic and mixed collaborations increased in absolute numbers in the last seven years of the review period (2013-2019). Regarding modelling practices between the two collaboration types, mixed collaborations were more likely to: (i) include authors who were affiliated to an institutions in the location studied, (ii) use more complex modelling practices including stochastic model dynamics, parametrization methods involving a combination of literature sources, expert opinion/assumptions, and fitting to data, and validate their models with data collected independent of the study. Even though this review was about human vaccine-preventable diseases, only 52.5\% of the studies modelled vaccination in some form. Influenza was disproportionately the most studied disease, followed by Ebola, and Dengue. Final epidemic size and attack rate were a commonly preferred modelling outcome of interest. Lastly, even though all the studies conducted simulation studies, code availability was rare as only $1.8\%$ of the studies $(4/227)$ made them readily accessible.

There were several differences in models and modelling practices/approaches between the human disease literature and foot and mouth disease (FMD). Mixed collaborations dominated the FMD literature and almost all the papers had an author in the location studied. Furthermore, most of the FMD models and modelling practices were more complex involving the use of spatial, stochastic, agent-based models. However, the FMD models were less likely to be validated with real world data.

We quantified the presence of mixed collaborations in outbreak response modelling between $1970$ and $2019$ (Fig \ref{studies_per_collab_type}). Our results show an absolute increase in the number of papers published by mixed collaborations during the period. Mixed or interdisciplinary collaborations have been reported to be strongly associated with research impact and translation to decision-making \cite{Deelstra2003}. We, however, did not measure the impact of mixed collaborations in outbreak response decision-making and therefore recommend that future studies consider pursuing this. Outbreak response, which is broadly an operational field, should ideally have modelling groups/collaborations that impact decision-making.

The increase in mixed collaborations could suggest an increase in the uptake or recognition of modelling by decision-makers in governmental and non-governmental organizations as an outbreak response decision-making tool. There are examples of the commissioning of modelling by organisations such as the US Centers for Disease Control and Prevention (CDC) and the World Health Organization (WHO) to support decision-making during past outbreaks of Influenza, Ebola, Zika, and Dengue \cite{Muscatello2017}. However, research that measures the direct impact and contributions of such collaborations to decision-making remains lacking in the literature and calls for future research \cite{Muscatello2017}.

The inclusion of local experts in outbreak response modelling teams helps with more tailored problem-solving and decision-making and better reception of the decisions made \cite{Abramowitz2015}. In that regard, mathematical modelling collaborations involving local experts or, in this review, locally affiliated co-authors are an important first step towards achieving this. We found that most of the human disease studies had authors with an institutional affiliation in the location studied. We observed the same pattern for FMD albeit at a much higher percentage. When the human disease data was disaggregated, the results showed that mixed collaborations had a much higher percentage.

### 2023-09-02 manuscript/submission2/outbreak_response_models_review_v2.tex

We found some commonalities and differences in the choice of model structure and dynamics between the two collaboration types in the human disease literature. Mixed collaborations were more likely to use agent-based model structures and model validation methods like comparing model outputs to observed data. These are often difficult to implement in practice and tend to lead to complex outcomes. We are not making any value judgements with regards to complexity but are only highlighting these differences that might require further investigation. Moreover, it is common to model an outbreak using alternative model choices and assumptions depending on the question being answered \cite{Basu2013}. However, the results must always be interpreted with cognisance of the limitations/assumptions of the model.  The use of more complex models and practices by mixed collaborations is likely due to the fact that often, real world public health policy-related decision-making is operational, requiring finer details in models and approaches to answer the questions posed in a practical way. Our definition of mixed collaborations meant that it may have involved decision-makers. It is, therefore, likely that their need for practical solutions could have influenced higher levels of model complexity.

The FMD models were generally more complex than the human disease models. This is not surprising because the nature of FMD spread often requires the inclusion of farm structure, farm connectivity, and demographics to capture the disease's dynamics accurately and to capture the complex nature of the responses to outbreaks in terms of movement control and depopulations of farms \cite{Kinsley2018}. These levels of detail and complexity are often required in human disease outbreak models, but we observed a more frequent use among the FMD studies. Future studies designed to explain the differences in modelling practices of human outbreak response modelling groups might help to explain what we have observed in this review.

A little over half of the human disease papers used observed data for parametrization or validation. Additionally, only 4 studies shared their code in an easy to access form. Owing to this, it might be difficult to reproduce some of the results in these outbreak response modelling studies. We recognise that data sharing in public health raises a lot of debate regarding privacy and intellectual property, and often, authors are hindered by institutional data sharing policies \cite{Kim2016}. We, however, recommend that data and code be shared where possible to promote open science practices that help advance the field.  The FMD models were generally not validated with observed data. This is most likely due to the lack of data on FMD outbreaks as most areas in the world are free of FMD outbreaks. The lack of data seems to have constrained FMD outbreak model validation and might serve as an opportunity to use data-free model validation techniques. However, these techniques were not captured in this review.

This review had several limitations. First, outbreak response models and analyses are not always published in peer-reviewed journals, but this systematic review only focused on peer-reviewed articles. It is, therefore, possible that some relevant studies were missed by our search strategy. Second, we used the author list and affiliations to classify studies as either purely academic or mixed. However, this could cause some mixed collaborations to be misclassified as purely academic, especially in cases where non-academics contributed to papers classified as purely academic but were not included on the author list. It is, however, standard practice to include individuals who contributed substantially to a piece of scientific writing, hence, if that was not done for a paper, it could imply that the level of interaction was not high enough to warrant the credit of authorship. Third, in the absence of an explicit measure of contribution of mathematical modelling to outbreak response decision-making, we used mixed collaborations as a proxy. Thus, we may be under-estimating the number of mixed collaborations in the literature. Fourth, we only surveyed the literature on a specific study design – mechanistic models. We excluded statistical models and other computational models, which are not mechanistic. Hence, the results of this systematic review should only be interpreted in the context of the mechanistic modelling landscape. Furthermore, it is also possible that some purely academic collaborations contribute to decision-making whereas some mixed collaborations are purely an academic exercise. Also, we  conducted our database searches in January 2020 and therefore do not reflect the literature, or any changes in practice, associated with the COVID-19 pandemic. This study, however, provides a baseline for comparing practices before and after the COVID-19 pandemic. Lastly, by only including papers published in English, we most certainly missed papers published in other languages.

@@

\paragraph*{S1 Fig}
\label{S1_Fig}
{\bf Collaboration patterns in time (proportions per year).}

\paragraph*{S2 File}
\label{S2_File}
{\bf Analysis of extracted data.}

\paragraph*{S3 Table}
\label{S3_Table}
{\bf PRISMA checklist.}

\paragraph*{S4 Table}
\label{S4_Table}
{\bf Database of included studies and extracted data per paper.}
